---
title: "Introduction to altmem"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

Install altmem.

```{r install, eval=FALSE}
BiocManager::install("igordot/altmem")
```

## Usage

Load altmem.

```{r setup}
library(altmem)
```

Load an example cytoMEM dataset from the mass cytometry characterization of normal human blood.

```{r data}
data(PBMC, package = "cytoMEM")
str(PBMC)
```

```{r}
head(PBMC)
```

Seven cell populations were identified (the labels are stored in the `cluster` column):

- 1: CD4+ T cells
- 2: CD8+ T cells
- 3: dendritic cells (DCs)
- 4: IgM- B cells
- 5: IgM+ B cells
- 6: monocytes
- 7: natural killer (NK) cells

This dataset is not transformed, so we will apply an arcsinh transformation with a cofactor of 15.

```{r data-asinh}
PBMCt <- PBMC[, colnames(PBMC) != "cluster"]
PBMCt <- asinh(PBMCt / 15)
PBMCt <- cbind(PBMCt, cluster = PBMC[, "cluster"])
```

Run marker enrichment modeling.

```{r altmem}
res <- altmem(PBMCt, cluster_col = "cluster")
```

Alternatively, the transformation can be done within the MEM function.

```{r altmem-transform, eval=FALSE}
res <- altmem(PBMC, cluster_col = "cluster", transform = TRUE, cofactor = 15)
```

The output is a list of matrices containing the MEM scores and related statistics.

```{r}
str(res)
```

After the labels are calculated, we can visualize the results.

```{r heatmap, fig.width=10}
mem_heatmap(res)
```

The heatmap can be further customized: `show_mem_labels` adds the descriptive MEM labels to the cluster names,`max_label_markers` limits the number of markers shown in the labels, and `min_heatmap_score` adds a threshold for keeping only variable markers.

```{r heatmap-labels, fig.width=10}
mem_heatmap(res, show_mem_labels = TRUE, max_label_markers = 3, min_heatmap_score = 2)
```

The MEM scores and labels agree with the known biology of these cell populations. For example, cluster 1 (CD4+ T cells) is enriched for CD4 and CD3, while cluster 2 (CD8+ T cells) is enriched for CD8 and CD3.

The labels can be extracted as a simple vector.

```{r labels}
mem_labels(res, min_label_score = 1, max_label_markers = 5, show_label_scores = FALSE)
```

## cytoMEM workflow

For comparison, we can run the same analysis using the original cytoMEM package.

```{r cytomem-mem}
cytomem_res <- cytoMEM::MEM(PBMC, transform = TRUE, cofactor = 15)
```

As expected, the initial `MEM()` output is the same as `altmem()`.

```{r}
identical(res, cytomem_res)
```

We can also generate the cytoMEM heatmap.

```{r cytomem-heatmap, fig.width=8}
cytoMEM::build_heatmaps(cytomem_res, only.MEMheatmap = TRUE)
```

```{r cytomem-heatmap-labels, fig.width=8}
cytoMEM::build_heatmaps(cytomem_res, display.thresh = 2, labels = TRUE, only.MEMheatmap = TRUE)
```
